---
title: "NYCSquirrelAnalysis"
author: "Ainsley Owens"
format: html
editor: visual
---

## Squirrels Analysis

This is an analysis of the NYC Squirrel Census dataset.

## Loading Packages and Datasets.

Here we loaded the tidymodels package and squirrels data.

```{r}
nyc_squirrels <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-29/nyc_squirrels.csv")

#install.packages("tidymodels")
library(tidymodels)
#Load the tidyverse
library(tidyverse)
library (kableExtra)

nyc_squirrels_splits <- initial_split(nyc_squirrels, prop = 0.5)

exploratory_data <- training(my_data_splits)
test_data <- testing(my_data_splits)
exploratory_data %>% 
  head(6)%>% 
  kable() %>% 
  kable_styling (c("striped", "hover"))
```

This table represents the geographical, physical, and behavioral features of the squirrels in the dataset.

## About Our Data

## Interesting Questions to Ask

-   What is the most common fur color?

-   How many squirrels of each fur color are there?

-   How many juveniles and adults are there?

-   Are squirrels more likely to be above ground or on the ground plane?

-   Are squirrels seen more at night or during the day?

-   What is the most commonly seen behavior?

-   Does most commonly seen behavior differ from juveniles and adults?

## Hypotheses

-   Most common fur color is gray.

-   There will be more adults than juveniles.

-   Squirrels are more likely to be above ground than on the ground plane.

-   Squirrels are more likely to be seen at night.

-   Most commonly seen behavior is running or eating.

## Answering Our Questions

What is the most common fur color? How many squirrels of each fur color are there?

```{r}
exploratory_data %>%
  select(primary_fur_color)
exploratory_data %>%
  count(primary_fur_color)
```

How many juveniles and adults are there?

```{r}
exploratory_data %>% 
  count (age)
```

Are squirrels more likely to be above ground or on the ground plane?

```{r}
exploratory_data %>% 
  count(location)
```

Are squirrels seen more at night or during the day?

```{r}
exploratory_data %>%
  count (shift)
```

What is the most commonly seen behavior?

```{r}
exploratory_data %>% 

```
